---
title: "WHOOP Decoded: A Year of Recovery Data"
format:
  closeread-html:
    css: styles.css
    cr-style:
      narrative-background-color: "#0d0d0d"
      narrative-text-color: "#e0e0e0"
      sticky-background-color: "#111111"
      narrative-overlay: false
---
```{r}
#| label: setup
#| include: false

library(tidyverse)
library(plotly)
library(lubridate)

main_df        <- readRDS("data/main_df.rds")
workouts_clean <- readRDS("data/workouts_clean.rds")

n_days       <- nrow(main_df)
avg_hrv      <- round(mean(main_df$hrv, na.rm = TRUE))
avg_recovery <- round(mean(main_df$recovery, na.rm = TRUE))
n_workouts   <- nrow(workouts_clean)
date_start   <- format(min(main_df$date), "%B %Y")
date_end     <- format(max(main_df$date), "%B %Y")
pct_green    <- round(mean(main_df$recovery_zone == "Green", na.rm = TRUE) * 100)

zone_colors <- c("Red" = "#FF6B6B", "Yellow" = "#FFD166", "Green" = "#00f19f")
marker_cols <- zone_colors[as.character(main_df$recovery_zone)]
```

::: hero
# WHOOP Decoded
### What does your recovery data actually tell you?


::: stats-row
::: stat-box
::: number
`r n_days`
:::
::: label
Days Tracked
:::
:::

::: stat-box
::: number
`r avg_hrv` ms
:::
::: label
Avg HRV
:::
:::

::: stat-box
::: number
`r avg_recovery`%
:::
::: label
Avg Recovery
:::
:::

::: stat-box
::: number
`r pct_green`%
:::
::: label
Days in the Green
:::
:::

::: stat-box
::: number
`r n_workouts`
:::
::: label
Workouts Logged
:::
:::
:::
:::

---

:::{.cr-section}

### Chart 1 · Your HRV Over a Year {.cr-narrative}

HRV — heart rate variability is time of variation between heartbeats in milliseconds. A high HRV means your body is calm and ready for excercise. A low HRV means your body is still dealing with stress, hard training, or poor sleep. [@cr-hrv]

::: insight
**Why it matters:** WHOOP gives you a recovery number, but HRV is the significant figure underneath it. Watching the trend over months reveals your fitness habits.
:::

Use the **range slider** at the bottom to zoom into any stretch of weeks. Hover any dot to see the full breakdown for that day. [@cr-hrv]

Patterns indicate consistencies or inconsistencies in training and recovery. Dips can be seen after harder training blocks or busier times of the year. Spikes can be seen after rest. [@cr-hrv]

:::{#cr-hrv}
```{r}
#| echo: false
#| warning: false

plot_ly(
  data   = main_df,
  x      = ~date,
  y      = ~hrv,
  type   = "scatter",
  mode   = "lines+markers",
  color  = ~recovery_zone,
  colors = zone_colors,
  marker = list(size = 6, opacity = 0.85),
  line   = list(width = 1.4),
  text   = ~paste0(
    "<b>", format(date, "%b %d, %Y"), "</b><br>",
    "HRV: ", hrv, " ms<br>",
    "Recovery: ", recovery, "%<br>",
    "Resting HR: ", rhr, " bpm<br>",
    "Sleep: ", sleep_perf, "%"
  ),
  hoverinfo = "text"
) |>
  layout(
    paper_bgcolor = "#111111",
    plot_bgcolor  = "#1a1a1a",
    font  = list(color = "#cccccc"),
    title = list(
      text = "HRV Over Time — One Dot Per Day",
      font = list(size = 16, color = "#ffffff")
    ),
    xaxis = list(
      title      = "",
      color      = "#888",
      gridcolor  = "#222",
      rangeslider = list(visible = TRUE, thickness = 0.06,
                         bgcolor = "#1a1a1a")
    ),
    yaxis = list(
      title     = "HRV (ms)",
      color     = "#888",
      gridcolor = "#222"
    ),
    legend = list(
      title      = list(text = "Zone", font = list(color = "#888")),
      font       = list(color = "#ccc"),
      bgcolor    = "#1a1a1a",
      bordercolor = "#333"
    ),
    hovermode = "closest"
  )
```
:::

:::

---

:::{.cr-section}

### Chart 2 · Which sleep metric actually predicts recovery? {.cr-narrative}

WHOOP says sleep drives recovery. But which part of sleep? Total hours? Efficiency? How much deep or REM sleep you got?

Use the dropdown to swap the x-axis between different sleep metrics. A tight upward cluster means that metric predicts your recovery. A scattered blob means it doesn't. [@cr-sleep]

::: insight
**Common finding:** Sleep performance % and sleep debt tend to be stronger predictors than raw hours — it's not just about quantity.
:::

Each dot is one day. The color still shows your recovery zone — green dots should cluster toward the top right. [@cr-sleep]

:::{#cr-sleep}
```{r}
#| echo: false
#| warning: false

x_options <- list(
  "Sleep Performance %" = main_df$sleep_perf,
  "Sleep Efficiency %"  = main_df$sleep_eff,
  "Hours Asleep"        = main_df$asleep_hrs,
  "Deep Sleep %"        = main_df$deep_pct,
  "REM %"               = main_df$rem_pct,
  "Sleep Debt (hrs)"    = main_df$sleep_debt_hrs
)
x_names <- names(x_options)

sleep_buttons <- lapply(seq_along(x_options), function(i) {
  list(
    method = "update",
    args   = list(
      list(x = list(x_options[[i]])),
      list(xaxis = list(title = x_names[i],
                        color = "#888", gridcolor = "#222"))
    ),
    label = x_names[i]
  )
})

plot_ly(
  data      = main_df,
  x         = ~sleep_perf,
  y         = ~recovery,
  type      = "scatter",
  mode      = "markers",
  marker    = list(
    color   = marker_cols,
    size    = 9,
    opacity = 0.78,
    line    = list(width = 0.8, color = "#0d0d0d")
  ),
  text = ~paste0(
    "<b>", format(date, "%b %d, %Y"), "</b><br>",
    "Sleep Perf: ", sleep_perf, "%<br>",
    "Recovery: ", recovery, "%<br>",
    "Hours Asleep: ", round(asleep_hrs, 1), "<br>",
    "Deep Sleep: ", round(deep_pct, 1), "%<br>",
    "Sleep Debt: ", round(sleep_debt_hrs, 1), " hrs"
  ),
  hoverinfo = "text"
) |>
  layout(
    paper_bgcolor = "#111111",
    plot_bgcolor  = "#1a1a1a",
    font  = list(color = "#cccccc"),
    title = list(
      text = "Sleep Quality → Next-Day Recovery",
      font = list(size = 16, color = "#ffffff")
    ),
    xaxis = list(title = "Sleep Performance %",
                 color = "#888", gridcolor = "#222"),
    yaxis = list(title = "Recovery Score (%)",
                 color = "#888", gridcolor = "#222"),
    updatemenus = list(list(
      buttons    = sleep_buttons,
      direction  = "down",
      showactive = TRUE,
      bgcolor    = "#1a1a1a",
      bordercolor = "#333",
      font       = list(color = "#ccc"),
      x = 0.0, xanchor = "left",
      y = 1.18, yanchor = "top"
    )),
    annotations = list(list(
      text = "Switch x-axis:",
      x = 0.0, xref = "paper",
      y = 1.23, yref = "paper",
      showarrow = FALSE,
      font = list(size = 11, color = "#888")
    ))
  )
```
:::

:::

---

:::{.cr-section}

### Chart 3 · The insight WHOOP never surfaces {.cr-narrative}

Here's what WHOOP doesn't tell you directly: **your HRV that morning predicts how hard a workout will feel.** When HRV is low, your heart beats faster to produce the same output — your "easy" run suddenly feels hard. [@cr-workout]

::: insight
**What to look for:** A downward trend — higher HRV on the x-axis should mean lower average heart rate on the y-axis. If the pattern holds for you, low-HRV days are days to go easy.
:::

Use the **activity buttons** to filter by sport. Running and Jiu Jitsu are the most revealing — cardio sports where heart rate directly measures effort. [@cr-workout]

:::{#cr-workout}
```{r}
#| echo: false
#| warning: false

workout_df <- main_df |>
  filter(!is.na(avg_hr), !is.na(hrv), !is.na(activity_clean))

activities <- sort(unique(workout_df$activity_clean))
n_acts     <- length(activities)

activity_palette <- c(
  "Running"          = "#00f19f",
  "Weightlifting"    = "#FF6B6B",
  "Powerlifting"     = "#FF9F45",
  "Jiu Jitsu"        = "#A78BFA",
  "Walking"          = "#60A5FA",
  "Basketball"       = "#F472B6",
  "Dance"            = "#FBBF24",
  "Cycling"          = "#34D399",
  "General Activity" = "#94A3B8",
  "Other"            = "#64748B"
)

p_workout <- plot_ly()

for (act in activities) {
  df_act  <- workout_df |> filter(activity_clean == act)
  act_col <- if (act %in% names(activity_palette))
               activity_palette[act] else "#94A3B8"

  p_workout <- p_workout |>
    add_trace(
      data   = df_act,
      x      = ~hrv,
      y      = ~avg_hr,
      type   = "scatter",
      mode   = "markers",
      name   = act,
      marker = list(size = 10, opacity = 0.78, color = act_col,
                    line = list(width = 0.5, color = "#0d0d0d")),
      text   = ~paste0(
        "<b>", format(date, "%b %d, %Y"), "</b><br>",
        "Activity: ", activity_clean, "<br>",
        "HRV: ", hrv, " ms<br>",
        "Avg HR: ", avg_hr, " bpm<br>",
        "Strain: ", round(strain, 1), "<br>",
        "Duration: ", round(duration_min), " min"
      ),
      hoverinfo = "text"
    )
}

all_btn <- list(
  method = "restyle",
  args   = list("visible", as.list(rep(TRUE, n_acts))),
  label  = "All"
)
activity_btns <- lapply(seq_along(activities), function(i) {
  vis    <- rep(FALSE, n_acts)
  vis[i] <- TRUE
  list(method = "restyle",
       args   = list("visible", as.list(vis)),
       label  = activities[i])
})

p_workout |>
  layout(
    paper_bgcolor = "#111111",
    plot_bgcolor  = "#1a1a1a",
    font  = list(color = "#cccccc"),
    title = list(
      text = "Morning HRV vs. Workout Heart Rate",
      font = list(size = 16, color = "#ffffff")
    ),
    xaxis = list(title = "HRV That Morning (ms)",
                 color = "#888", gridcolor = "#222"),
    yaxis = list(title = "Avg Workout Heart Rate (bpm)",
                 color = "#888", gridcolor = "#222"),
    updatemenus = list(list(
      type       = "buttons",
      buttons    = c(list(all_btn), activity_btns),
      direction  = "right",
      showactive = TRUE,
      bgcolor    = "#1a1a1a",
      bordercolor = "#333",
      font       = list(color = "#ccc", size = 11),
      x = 0.0, xanchor = "left",
      y = 1.14, yanchor = "top"
    )),
    legend    = list(font = list(color = "#ccc"),
                     bgcolor = "#1a1a1a", bordercolor = "#333"),
    hovermode = "closest"
  )
```
:::

:::

---

:::{.cr-section}

### Chart 4 · Discover Trends {.cr-narrative}

Pick any variable for X and Y — every dot is one day, colored by recovery zone. [@cr-explore]



:::{#cr-explore}
```{r}
#| echo: false
#| warning: false

explore_vars <- list(
  "Recovery (%)"          = main_df$recovery,
  "HRV (ms)"              = main_df$hrv,
  "Resting HR (bpm)"      = main_df$rhr,
  "Sleep Performance (%)" = main_df$sleep_perf,
  "Sleep Efficiency (%)"  = main_df$sleep_eff,
  "Hours Asleep"          = main_df$asleep_hrs,
  "Deep Sleep (%)"        = main_df$deep_pct,
  "REM (%)"               = main_df$rem_pct,
  "Sleep Debt (hrs)"      = main_df$sleep_debt_hrs,
  "Day Strain"            = main_df$day_strain,
  "Resp. Rate (rpm)"      = main_df$resp_rate,
  "Workout Avg HR (bpm)"  = main_df$avg_hr
)
var_names <- names(explore_vars)

make_btns <- function(axis) {
  lapply(seq_along(explore_vars), function(i) {
    args <- if (axis == "x")
      list(list(x = list(explore_vars[[i]])),
           list(xaxis = list(title = var_names[i],
                             color = "#888", gridcolor = "#222")))
    else
      list(list(y = list(explore_vars[[i]])),
           list(yaxis = list(title = var_names[i],
                             color = "#888", gridcolor = "#222")))
    list(method = "update", args = args, label = var_names[i])
  })
}

plot_ly(
  data      = main_df,
  x         = ~hrv,
  y         = ~recovery,
  type      = "scatter",
  mode      = "markers",
  marker    = list(color = marker_cols, size = 9, opacity = 0.75,
                   line = list(width = 0.8, color = "#0d0d0d")),
  text      = ~paste0(
    "<b>", format(date, "%b %d, %Y"), "</b><br>",
    "Recovery: ", recovery, "%<br>",
    "HRV: ", hrv, " ms<br>",
    "Sleep: ", sleep_perf, "%<br>",
    "Day Strain: ", round(day_strain, 1)
  ),
  hoverinfo = "text"
) |>
  layout(
    paper_bgcolor = "#111111",
    plot_bgcolor  = "#1a1a1a",
    font  = list(color = "#cccccc"),
    title = list(
      text = "Explore Your Data — Pick Any X & Y",
      font = list(size = 16, color = "#ffffff")
    ),
    xaxis = list(title = "HRV (ms)", color = "#888", gridcolor = "#222"),
    yaxis = list(title = "Recovery (%)", color = "#888", gridcolor = "#222"),
    updatemenus = list(
      list(buttons = make_btns("x"), direction = "down",
           showactive = TRUE, bgcolor = "#1a1a1a",
           bordercolor = "#333", font = list(color = "#ccc"),
           x = 0.0, xanchor = "left", y = 1.22, yanchor = "top"),
      list(buttons = make_btns("y"), direction = "down",
           showactive = TRUE, bgcolor = "#1a1a1a",
           bordercolor = "#333", font = list(color = "#ccc"),
           x = 0.45, xanchor = "left", y = 1.22, yanchor = "top")
    ),
    annotations = list(
      list(text = "X axis:", x = 0.0, xref = "paper",
           y = 1.27, yref = "paper", showarrow = FALSE,
           font = list(size = 11, color = "#888")),
      list(text = "Y axis:", x = 0.45, xref = "paper",
           y = 1.27, yref = "paper", showarrow = FALSE,
           font = list(size = 11, color = "#888"))
    )
  )
```
:::

:::

---

:::{.cr-section}

### Chart 5 · 365 Days at a Glance {.cr-narrative}

Every square is one day. Green means your body was ready. Red means it wasn't. Yellow is in between. [@cr-calendar]

::: insight
**What to look for:** Streaks of red often follow heavy training blocks or poor sleep weeks. Green clusters usually mean you backed off at the right time.
:::

Scroll back through the year — patterns in the calendar reveal things the line chart buries. A bad month looks very different from a bad week. Hover any square to see the full breakdown for that day. [@cr-calendar]

:::{#cr-calendar}
```{r}
#| echo: false
#| warning: false

start_week <- floor_date(min(main_df$date), 'week', week_start = 1)

cal_df <- main_df |>
  mutate(
    dow        = as.integer(wday(date, week_start = 1)) - 1,
    week_num   = as.integer(floor(as.numeric(date - start_week) / 7)),
    zone_color = zone_colors[as.character(recovery_zone)],
    hover_text = paste0(
      "<b>", format(date, "%b %d, %Y"), "</b><br>",
      "Recovery: ", recovery, "%<br>",
      "Zone: ", as.character(recovery_zone), "<br>",
      "HRV: ", hrv, " ms<br>",
      "Resting HR: ", rhr, " bpm"
    )
  )

month_ticks <- cal_df |>
  mutate(month_label = format(date, "%b '%y")) |>
  group_by(month_label) |>
  slice_min(date, n = 1) |>
  ungroup()

plot_ly(
  data      = cal_df,
  x         = ~week_num,
  y         = ~dow,
  type      = "scatter",
  mode      = "markers",
  marker    = list(
    symbol = "square",
    size   = 16,
    color  = ~zone_color,
    line   = list(width = 2, color = "#0d0d0d")
  ),
  text      = ~hover_text,
  hoverinfo = "text"
) |>
  layout(
    paper_bgcolor = "#111111",
    plot_bgcolor  = "#111111",
    font  = list(color = "#cccccc"),
    title = list(
      text = "Recovery Calendar — One Square Per Day",
      font = list(size = 16, color = "#ffffff")
    ),
    xaxis = list(
      title    = "",
      tickvals = month_ticks$week_num,
      ticktext = month_ticks$month_label,
      color    = "#888",
      showgrid = FALSE,
      zeroline = FALSE
    ),
    yaxis = list(
      title     = "",
      tickvals  = 0:6,
      ticktext  = c("Mon","Tue","Wed","Thu","Fri","Sat","Sun"),
      color     = "#888",
      showgrid  = FALSE,
      zeroline  = FALSE,
      autorange = "reversed"
    ),
    showlegend  = FALSE,
    hovermode   = "closest",
    annotations = list(
      list(text = "■ Green", x = 0.0, xref = "paper",
           y = -0.15, yref = "paper", showarrow = FALSE,
           font = list(size = 13, color = "#00f19f")),
      list(text = "■ Yellow", x = 0.15, xref = "paper",
           y = -0.15, yref = "paper", showarrow = FALSE,
           font = list(size = 13, color = "#FFD166")),
      list(text = "■ Red", x = 0.30, xref = "paper",
           y = -0.15, yref = "paper", showarrow = FALSE,
           font = list(size = 13, color = "#FF6B6B"))
    )
  )
```
:::

:::
