---
title: "WHOOP Decoded: A Year of Recovery Data"
format:
  closeread-html:
    css: styles.css
    cr-style:
      narrative-background-color: "#0d0d0d"
      narrative-text-color: "#e0e0e0"
      sticky-background-color: "#111111"
      narrative-overlay: false
---
```{r}
#| label: setup
#| include: false

library(tidyverse)
library(plotly)
library(lubridate)

main_df        <- readRDS("data/main_df.rds")
workouts_clean <- readRDS("data/workouts_clean.rds")

n_days       <- nrow(main_df)
avg_hrv      <- round(mean(main_df$hrv, na.rm = TRUE))
avg_recovery <- round(mean(main_df$recovery, na.rm = TRUE))
n_workouts   <- nrow(workouts_clean)
date_start   <- format(min(main_df$date), "%B %Y")
date_end     <- format(max(main_df$date), "%B %Y")
pct_green    <- round(mean(main_df$recovery_zone == "Green", na.rm = TRUE) * 100)

zone_colors <- c("Red" = "#FF6B6B", "Yellow" = "#FFD166", "Green" = "#00f19f")
marker_cols <- zone_colors[as.character(main_df$recovery_zone)]
```

::: hero
# WHOOP Decoded
### What does your recovery data actually tell you?


::: stats-row
::: stat-box
::: number
`r n_days`
:::
::: label
Days Tracked
:::
:::

::: stat-box
::: number
`r avg_hrv` ms
:::
::: label
Avg HRV
:::
:::

::: stat-box
::: number
`r avg_recovery`%
:::
::: label
Avg Recovery
:::
:::

::: stat-box
::: number
`r pct_green`%
:::
::: label
Days in the Green
:::
:::

::: stat-box
::: number
`r n_workouts`
:::
::: label
Workouts Logged
:::
:::
:::
:::

---

:::{.cr-section}

### Chart 1 · Your HRV Over a Year {.cr-narrative}

HRV — heart rate variability is the variation between heartbeats in milliseconds. A high HRV means your body is calm and ready for exercise. A low HRV means your body is still dealing with stress, hard training, or poor sleep. [@cr-hrv]

::: insight
**Why it matters:** WHOOP gives you a recovery number, but HRV is the significant figure underneath it. Watching the trend over months reveals your fitness habits.
:::

Use the **range slider** at the bottom to zoom into any stretch of weeks. Hover any dot to see the full breakdown for that day. The white line shows your 7-day rolling average — the true signal beneath the daily noise. [@cr-hrv]

Patterns indicate consistencies or inconsistencies in training and recovery. Dips can be seen after harder training blocks or busier times of the year. Spikes can be seen after rest. [@cr-hrv]

:::{#cr-hrv}
```{r}
#| echo: false
#| warning: false

main_df <- main_df |>
  arrange(date) |>
  mutate(hrv_7day = zoo::rollmean(hrv, k = 7, fill = NA, align = "right"))

p_hrv <- plot_ly(
  data   = main_df,
  x      = ~date,
  y      = ~hrv,
  type   = "scatter",
  mode   = "lines+markers",
  color  = ~recovery_zone,
  colors = zone_colors,
  marker = list(size = 6, opacity = 0.85),
  line   = list(width = 1.4),
  text   = ~paste0(
    "<b>", format(date, "%b %d, %Y"), "</b><br>",
    "HRV: ", hrv, " ms<br>",
    "Recovery: ", recovery, "%<br>",
    "Resting HR: ", rhr, " bpm<br>",
    "Sleep: ", sleep_perf, "%"
  ),
  hoverinfo = "text"
)

p_hrv <- p_hrv |>
  add_trace(
    data       = main_df,
    x          = ~date,
    y          = ~hrv_7day,
    type       = "scatter",
    mode       = "lines",
    name       = "7-day avg",
    line       = list(color = "#ffffff", width = 2.5, dash = "solid"),
    hoverinfo  = "skip",
    inherit    = FALSE
  )

p_hrv |>
  layout(
    paper_bgcolor = "#111111",
    plot_bgcolor  = "#1a1a1a",
    font  = list(color = "#cccccc"),
    title = list(
      text = "HRV Over Time — One Dot Per Day",
      font = list(size = 16, color = "#ffffff")
    ),
    xaxis = list(
      title      = "",
      color      = "#888",
      gridcolor  = "#222",
      rangeslider = list(visible = TRUE, thickness = 0.06,
                         bgcolor = "#1a1a1a")
    ),
    yaxis = list(
      title     = "HRV (ms)",
      color     = "#888",
      gridcolor = "#222"
    ),
    legend = list(
      title      = list(text = "Zone", font = list(color = "#888")),
      font       = list(color = "#ccc"),
      bgcolor    = "#1a1a1a",
      bordercolor = "#333"
    ),
    hovermode = "closest"
  )
```
:::

:::

---

:::{.cr-section}

### Chart 2 · Which sleep metric actually predicts recovery? {.cr-narrative}

WHOOP says sleep drives recovery. But which part of sleep? Total hours? Efficiency? How much deep or REM sleep you got?

Use the dropdown to swap the x-axis between different sleep metrics. A tight upward cluster means that metric predicts your recovery. A scattered blob means it does not. [@cr-sleep]

::: insight
**Common finding:** Sleep performance % and sleep debt tend to be stronger predictors than raw hours — it is not just about quantity.
:::

Each dot is one day. The color still shows your recovery zone — green dots should cluster toward the top right. [@cr-sleep]

:::{#cr-sleep}
```{r}
#| echo: false
#| warning: false

x_options <- list(
  "Sleep Performance %" = main_df$sleep_perf,
  "Sleep Efficiency %"  = main_df$sleep_eff,
  "Hours Asleep"        = main_df$asleep_hrs,
  "Deep Sleep %"        = main_df$deep_pct,
  "REM %"               = main_df$rem_pct,
  "Sleep Debt (hrs)"    = main_df$sleep_debt_hrs
)
x_names <- names(x_options)

sleep_buttons <- lapply(seq_along(x_options), function(i) {
  list(
    method = "update",
    args   = list(
      list(x = list(x_options[[i]])),
      list(xaxis = list(title = x_names[i],
                        color = "#888", gridcolor = "#222"))
    ),
    label = x_names[i]
  )
})

plot_ly(
  data      = main_df,
  x         = ~sleep_perf,
  y         = ~recovery,
  type      = "scatter",
  mode      = "markers",
  marker    = list(
    color   = marker_cols,
    size    = 9,
    opacity = 0.78,
    line    = list(width = 0.8, color = "#0d0d0d")
  ),
  text = ~paste0(
    "<b>", format(date, "%b %d, %Y"), "</b><br>",
    "Sleep Perf: ", sleep_perf, "%<br>",
    "Recovery: ", recovery, "%<br>",
    "Hours Asleep: ", round(asleep_hrs, 1), "<br>",
    "Deep Sleep: ", round(deep_pct, 1), "%<br>",
    "Sleep Debt: ", round(sleep_debt_hrs, 1), " hrs"
  ),
  hoverinfo = "text"
) |>
  layout(
    paper_bgcolor = "#111111",
    plot_bgcolor  = "#1a1a1a",
    font  = list(color = "#cccccc"),
    title = list(
      text = "Sleep Quality to Next-Day Recovery",
      font = list(size = 16, color = "#ffffff")
    ),
    xaxis = list(title = "Sleep Performance %",
                 color = "#888", gridcolor = "#222"),
    yaxis = list(title = "Recovery Score (%)",
                 color = "#888", gridcolor = "#222"),
    updatemenus = list(list(
      buttons    = sleep_buttons,
      direction  = "down",
      showactive = TRUE,
      bgcolor    = "#1a1a1a",
      bordercolor = "#333",
      font       = list(color = "#ccc"),
      x = 0.0, xanchor = "left",
      y = 1.18, yanchor = "top"
    )),
    annotations = list(list(
      text = "Switch x-axis:",
      x = 0.0, xref = "paper",
      y = 1.23, yref = "paper",
      showarrow = FALSE,
      font = list(size = 11, color = "#888")
    ))
  )
```
:::

:::

---

:::{.cr-section}

### Chart 3 · The insight WHOOP never surfaces {.cr-narrative}

Here is what WHOOP does not tell you directly: **your HRV that morning predicts how hard a workout will feel.** When HRV is low, your heart beats faster to produce the same output — your easy run suddenly feels hard. [@cr-workout]

::: insight
**What to look for:** A downward trend — higher HRV on the x-axis should mean lower average heart rate on the y-axis. If the pattern holds for you, low-HRV days are days to go easy.
:::

Use the **activity buttons** to filter by sport. Running and Jiu Jitsu are the most revealing — cardio sports where heart rate directly measures effort. On low-HRV days, the same pace requires a meaningfully higher heart rate. [@cr-workout]

The pattern is clearest in Running — filter to it and watch how the dots slope downward from left to right. Higher HRV mornings consistently produce lower workout heart rates at the same effort level. [@cr-workout]

:::{#cr-workout}
```{r}
#| echo: false
#| warning: false

workout_df <- main_df |>
  filter(!is.na(avg_hr), !is.na(hrv), !is.na(activity_clean))

activities <- sort(unique(workout_df$activity_clean))
n_acts     <- length(activities)

activity_palette <- c(
  "Running"          = "#00f19f",
  "Weightlifting"    = "#FF6B6B",
  "Powerlifting"     = "#FF9F45",
  "Jiu Jitsu"        = "#A78BFA",
  "Walking"          = "#60A5FA",
  "General Activity" = "#94A3B8",
  "Other"            = "#64748B"
)

p_workout <- plot_ly()

for (act in activities) {
  df_act  <- workout_df |> filter(activity_clean == act)
  act_col <- if (act %in% names(activity_palette))
               activity_palette[act] else "#94A3B8"

  p_workout <- p_workout |>
    add_trace(
      data   = df_act,
      x      = ~hrv,
      y      = ~avg_hr,
      type   = "scatter",
      mode   = "markers",
      name   = act,
      marker = list(size = 10, opacity = 0.78, color = act_col,
                    line = list(width = 0.5, color = "#0d0d0d")),
      text   = ~paste0(
        "<b>", format(date, "%b %d, %Y"), "</b><br>",
        "Activity: ", activity_clean, "<br>",
        "HRV: ", hrv, " ms<br>",
        "Avg HR: ", avg_hr, " bpm<br>",
        "Strain: ", round(strain, 1), "<br>",
        "Duration: ", round(duration_min), " min"
      ),
      hoverinfo = "text"
    )
}

all_btn <- list(
  method = "restyle",
  args   = list("visible", as.list(rep(TRUE, n_acts))),
  label  = "All"
)
activity_btns <- lapply(seq_along(activities), function(i) {
  vis    <- rep(FALSE, n_acts)
  vis[i] <- TRUE
  list(method = "restyle",
       args   = list("visible", as.list(vis)),
       label  = activities[i])
})

p_workout |>
  layout(
    paper_bgcolor = "#111111",
    plot_bgcolor  = "#1a1a1a",
    font  = list(color = "#cccccc"),
    title = list(
      text = "Morning HRV vs. Workout Heart Rate",
      font = list(size = 16, color = "#ffffff")
    ),
    xaxis = list(title = "HRV That Morning (ms)",
                 color = "#888", gridcolor = "#222"),
    yaxis = list(title = "Avg Workout Heart Rate (bpm)",
                 color = "#888", gridcolor = "#222"),
    updatemenus = list(list(
      type       = "buttons",
      buttons    = c(list(all_btn), activity_btns),
      direction  = "right",
      showactive = TRUE,
      bgcolor    = "#1a1a1a",
      bordercolor = "#333",
      font       = list(color = "#ccc", size = 11),
      x = 0.0, xanchor = "left",
      y = 1.14, yanchor = "top"
    )),
    legend    = list(font = list(color = "#ccc"),
                     bgcolor = "#1a1a1a", bordercolor = "#333"),
    hovermode = "closest"
  )
```
:::

:::

---

:::{.cr-section}

### Chart 4 · Discover Trends {.cr-narrative}

Pick any variable for X and Y — every dot is one day, colored by recovery zone.

Try **HRV vs. Resting HR** — these two move in opposite directions when your body is stressed. Or **Sleep Debt vs. Recovery** to see how cumulative sleep loss compounds over time. [@cr-explore]

::: insight
**Where to start:** Set X to Sleep Debt and Y to HRV. A downward slope means the longer your debt builds, the harder your nervous system works to recover.
:::

Green dots should cluster toward the top-right of any meaningful pairing. Scattered dots mean those two variables do not predict each other — that is useful information too. [@cr-explore]

:::{#cr-explore}
```{r}
#| echo: false
#| warning: false

explore_vars <- list(
  "Recovery (%)"          = main_df$recovery,
  "HRV (ms)"              = main_df$hrv,
  "Resting HR (bpm)"      = main_df$rhr,
  "Sleep Performance (%)" = main_df$sleep_perf,
  "Sleep Efficiency (%)"  = main_df$sleep_eff,
  "Hours Asleep"          = main_df$asleep_hrs,
  "Deep Sleep (%)"        = main_df$deep_pct,
  "REM (%)"               = main_df$rem_pct,
  "Sleep Debt (hrs)"      = main_df$sleep_debt_hrs,
  "Day Strain"            = main_df$day_strain,
  "Resp. Rate (rpm)"      = main_df$resp_rate,
  "Workout Avg HR (bpm)"  = main_df$avg_hr
)
var_names <- names(explore_vars)

make_btns <- function(axis) {
  lapply(seq_along(explore_vars), function(i) {
    args <- if (axis == "x")
      list(list(x = list(explore_vars[[i]])),
           list(xaxis = list(title = var_names[i],
                             color = "#888", gridcolor = "#222")))
    else
      list(list(y = list(explore_vars[[i]])),
           list(yaxis = list(title = var_names[i],
                             color = "#888", gridcolor = "#222")))
    list(method = "update", args = args, label = var_names[i])
  })
}

plot_ly(
  data      = main_df,
  x         = ~hrv,
  y         = ~recovery,
  type      = "scatter",
  mode      = "markers",
  marker    = list(color = marker_cols, size = 9, opacity = 0.75,
                   line = list(width = 0.8, color = "#0d0d0d")),
  text      = ~paste0(
    "<b>", format(date, "%b %d, %Y"), "</b><br>",
    "Recovery: ", recovery, "%<br>",
    "HRV: ", hrv, " ms<br>",
    "Sleep: ", sleep_perf, "%<br>",
    "Day Strain: ", round(day_strain, 1)
  ),
  hoverinfo = "text"
) |>
  layout(
    paper_bgcolor = "#111111",
    plot_bgcolor  = "#1a1a1a",
    font  = list(color = "#cccccc"),
    title = list(
      text = "Explore Your Data — Pick Any X and Y",
      font = list(size = 16, color = "#ffffff")
    ),
    xaxis = list(title = "HRV (ms)", color = "#888", gridcolor = "#222"),
    yaxis = list(title = "Recovery (%)", color = "#888", gridcolor = "#222"),
    updatemenus = list(
      list(buttons = make_btns("x"), direction = "down",
           showactive = TRUE, bgcolor = "#1a1a1a",
           bordercolor = "#333", font = list(color = "#ccc"),
           x = 0.0, xanchor = "left", y = 1.22, yanchor = "top"),
      list(buttons = make_btns("y"), direction = "down",
           showactive = TRUE, bgcolor = "#1a1a1a",
           bordercolor = "#333", font = list(color = "#ccc"),
           x = 0.45, xanchor = "left", y = 1.22, yanchor = "top")
    ),
    annotations = list(
      list(text = "X axis:", x = 0.0, xref = "paper",
           y = 1.27, yref = "paper", showarrow = FALSE,
           font = list(size = 11, color = "#888")),
      list(text = "Y axis:", x = 0.45, xref = "paper",
           y = 1.27, yref = "paper", showarrow = FALSE,
           font = list(size = 11, color = "#888"))
    )
  )
```
:::

:::

---

:::{.cr-section}

### Chart 5 · 365 Days at a Glance {.cr-narrative}

Every square is one day. Green means your body was ready. Red means it was not. Yellow is in between. [@cr-calendar]

::: insight
**What to look for:** Streaks of red often follow heavy training blocks or poor sleep weeks. Green clusters usually mean you backed off at the right time.
:::

Scroll back through the year — patterns in the calendar reveal things the line chart buries. A bad month looks very different from a bad week. Hover any square to see the full breakdown for that day. [@cr-calendar]

:::{#cr-calendar}
```{r}
#| echo: false
#| warning: false

start_week <- floor_date(min(main_df$date), "week", week_start = 1)

cal_df <- main_df |>
  mutate(
    dow        = as.integer(wday(date, week_start = 1)) - 1,
    week_num   = as.integer(floor(as.numeric(date - start_week) / 7)),
    zone_color = zone_colors[as.character(recovery_zone)],
    hover_text = paste0(
      "<b>", format(date, "%b %d, %Y"), "</b><br>",
      "Recovery: ", recovery, "%<br>",
      "Zone: ", as.character(recovery_zone), "<br>",
      "HRV: ", hrv, " ms<br>",
      "Resting HR: ", rhr, " bpm"
    )
  )

month_ticks <- cal_df |>
  mutate(month_label = format(date, "%b %y")) |>
  group_by(month_label) |>
  slice_min(date, n = 1) |>
  ungroup()

plot_ly(
  data      = cal_df,
  x         = ~week_num,
  y         = ~dow,
  type      = "scatter",
  mode      = "markers",
  marker    = list(
    symbol = "square",
    size   = 16,
    color  = ~zone_color,
    line   = list(width = 2, color = "#0d0d0d")
  ),
  text      = ~hover_text,
  hoverinfo = "text"
) |>
  layout(
    paper_bgcolor = "#111111",
    plot_bgcolor  = "#111111",
    font  = list(color = "#cccccc"),
    title = list(
      text = "Recovery Calendar — One Square Per Day",
      font = list(size = 16, color = "#ffffff")
    ),
    xaxis = list(
      title    = "",
      tickvals = month_ticks$week_num,
      ticktext = month_ticks$month_label,
      color    = "#888",
      showgrid = FALSE,
      zeroline = FALSE
    ),
    yaxis = list(
      title     = "",
      tickvals  = 0:6,
      ticktext  = c("Mon","Tue","Wed","Thu","Fri","Sat","Sun"),
      color     = "#888",
      showgrid  = FALSE,
      zeroline  = FALSE,
      autorange = "reversed"
    ),
    showlegend  = FALSE,
    hovermode   = "closest",
    annotations = list(
      list(text = "Green", x = 0.0, xref = "paper",
           y = -0.15, yref = "paper", showarrow = FALSE,
           font = list(size = 13, color = "#00f19f")),
      list(text = "Yellow", x = 0.12, xref = "paper",
           y = -0.15, yref = "paper", showarrow = FALSE,
           font = list(size = 13, color = "#FFD166")),
      list(text = "Red", x = 0.24, xref = "paper",
           y = -0.15, yref = "paper", showarrow = FALSE,
           font = list(size = 13, color = "#FF6B6B"))
    )
  )
```
:::

:::

---

:::{.cr-section}

### Chart 6 · What Do I Need Tonight? {.cr-narrative}

WHOOP tells you your recovery score after you wake up. This tool works the other way — enter what your body looked like today, and it tells you exactly how well you need to sleep tonight to wake up in the Green. [@cr-predict]

::: insight
**How it works:** A linear model trained on your 365 days of data solves backwards from your recovery target. Given your HRV and today's strain, it calculates the minimum sleep performance needed to hit each zone.
:::

Set your HRV from this morning and today's strain. Use this before bed to set a concrete sleep target — not a vague "sleep more" but an actual number. [@cr-predict]

This is what WHOOP never tells you. [@cr-predict]

:::{#cr-predict}
```{=html}
<div style="background:#1a1a1a;border-radius:16px;padding:36px 40px;max-width:620px;margin:0 auto;font-family:Inter,sans-serif;">
<div style="margin-bottom:28px;">
<label style="color:#888;font-size:0.75rem;text-transform:uppercase;letter-spacing:0.08em;">This Morning's HRV (ms)</label>
<div style="display:flex;align-items:center;gap:16px;margin-top:8px;">
<input type="range" id="hrv-slider" min="19" max="127" value="81" style="flex:1;accent-color:#00f19f;" oninput="calcSleep()">
<span id="hrv-val" style="color:#00f19f;font-size:1.4rem;font-weight:700;min-width:48px;text-align:right;">81</span>
</div></div>
<div style="margin-bottom:36px;">
<label style="color:#888;font-size:0.75rem;text-transform:uppercase;letter-spacing:0.08em;">Today's Strain</label>
<div style="display:flex;align-items:center;gap:16px;margin-top:8px;">
<input type="range" id="strain-slider" min="1" max="21" value="14" step="0.1" style="flex:1;accent-color:#00f19f;" oninput="calcSleep()">
<span id="strain-val" style="color:#00f19f;font-size:1.4rem;font-weight:700;min-width:48px;text-align:right;">14.0</span>
</div></div>
<div style="border-top:1px solid #2a2a2a;padding-top:28px;">
<div style="font-size:0.75rem;color:#666;text-transform:uppercase;letter-spacing:0.08em;margin-bottom:20px;text-align:center;">Sleep Performance Needed Tonight</div>
<div style="display:flex;flex-direction:column;gap:14px;">
<div style="display:flex;align-items:center;justify-content:space-between;background:#0d0d0d;border-radius:10px;padding:14px 20px;border-left:4px solid #00f19f;">
<span style="color:#00f19f;font-weight:600;font-size:0.95rem;">Green Zone (67%+ recovery)</span>
<span id="green-sleep" style="color:#00f19f;font-size:1.5rem;font-weight:800;">82%</span>
</div>
<div style="display:flex;align-items:center;justify-content:space-between;background:#0d0d0d;border-radius:10px;padding:14px 20px;border-left:4px solid #FFD166;">
<span style="color:#FFD166;font-weight:600;font-size:0.95rem;">Yellow Zone (34%+ recovery)</span>
<span id="yellow-sleep" style="color:#FFD166;font-size:1.5rem;font-weight:800;">38%</span>
</div>
<div style="display:flex;align-items:center;justify-content:space-between;background:#0d0d0d;border-radius:10px;padding:14px 20px;border-left:4px solid #FF6B6B;">
<span style="color:#FF6B6B;font-weight:600;font-size:0.95rem;">Avoid Red (stay above 34%)</span>
<span id="red-sleep" style="color:#FF6B6B;font-size:1.5rem;font-weight:800;">38%</span>
</div>
</div>
<div id="summary-text" style="margin-top:20px;text-align:center;font-size:0.9rem;color:#aaa;line-height:1.6;">
With HRV 81 and strain 14, aim for 82% sleep performance tonight to wake up Green.
</div>
</div></div>
<script>
function calcSleep(){
var hrv=parseFloat(document.getElementById("hrv-slider").value);
var strain=parseFloat(document.getElementById("strain-slider").value);
document.getElementById("hrv-val").textContent=Math.round(hrv);
document.getElementById("strain-val").textContent=strain.toFixed(1);
var intercept=-38.73;
var b_hrv=0.584;
var b_sleep=0.749;
var b_strain=-0.215;
var base=intercept+(b_hrv*hrv)+(b_strain*strain);
function sleepFor(target){
var s=(target-base)/b_sleep;
return Math.min(100,Math.max(9,Math.round(s)));
}
var greenSleep=sleepFor(67);
var yellowSleep=sleepFor(34);
document.getElementById("green-sleep").textContent=greenSleep+"%";
document.getElementById("yellow-sleep").textContent=yellowSleep+"%";
document.getElementById("red-sleep").textContent=yellowSleep+"%";
var summary="";
if(greenSleep<=100){
summary="With HRV "+Math.round(hrv)+" and strain "+strain.toFixed(1)+", aim for "+greenSleep+"% sleep performance tonight to wake up Green.";
}else{
summary="With HRV "+Math.round(hrv)+" and strain "+strain.toFixed(1)+", hitting Green tonight is unlikely — focus on at least "+yellowSleep+"% to stay in Yellow.";
}
document.getElementById("summary-text").textContent=summary;
}
calcSleep();
</script>
```
:::

:::